HO unification from object language to meta language

### 2024-02-27 paper/main.tex
4

\section{Languages description}
\label{sec:lang-spec}

@@

\begin{figure}[H]
  \begin{tabular}{ll}
  \begin{minipage}{0.21\textwidth}
  \begin{elpicodetab}
kind fm type.
type fapp list fm -> fm.
type flam (fm -> fm) -> fm.
type fcon string -> fm.
type fuva nat -> fm.
\end{elpicodetab}
  \end{minipage}
  &
  \begin{minipage}{0.24\textwidth}
  \begin{elpicodetab}
kind tm type.
type app list tm -> tm.
type lam (tm -> tm) -> tm.
type con string -> tm.
type uva nat -> list tm -> tm.
  \end{elpicodetab}
  \end{minipage}
  \end{tabular}
  \caption{\Fo{} and \Ho language}
  \label{code:common-terms}
  \Description[code:common-terms]{code:common-terms}
\end{figure}

\noindent
In the case of \Fo{} unification variables \elpiIn{fuva} have no explicit scope:
Unification variables standing for functions are applied to variables
via the \elpiIn{fapp} constructor.
For example in the statement of the instance \coqIn{forall_dec} the
term \coqIn{P x} is represented as \elpiIn{fapp[fuva N, x]}, where \elpiIn{N}
is of type \elpiIn{nat} and \elpiIn{x} is a bound variable.

@@

\noindent
The \elpiIn{name} builtin predicate tests if a term is a bound variable.

@@

\marginpar{maybe move code to the appendix and keep only signatures}
\begin{elpicode}
typeabbrev memory A (list (option A)).
type set nat -> memory A -> A -> o.
set N S T :- nth N S (some T).
type unset nat -> memory A -> o.
unset N S :- nth N S none.
type assign nat -> memory A -> A -> memory A -> o.
assign z [none|M] T [some T|M].
assign (s N) [X|M] T [X|M1] :- assign N M T M1. 
\end{elpicode}

\noindent 
If a memory cell is \elpiIn{none}, then the corresponding unification variable
is not set.

Since \Ho unification variables have a scope, their solution needs to be
abstracted over these in order to enables the instantiation of a single
solution to different scopes. On the contrary solutions to \Fo variables
are just terms.

\begin{elpicode}
  typeabbrev fsubst (memory ftm).

  kind assmt type.
  type abs (tm -> assmt) -> assmt.
  type val tm -> assmt.
  typeabbrev subst (memory assmt).
\end{elpicode}

\noindent
One can see \elpiIn{abs} as a syntax for the lambda abstraction of the meta
language.
%Luckily  we need to reason on these abstractions only when we
%manipulate the solution given by the unification of the meta language; when
%we manipulate the terms we don't 
%
%Another important implementation should be provided in order to represent the
%lambda abstractions and the applications of the ML. As outlined in
%\cref{sec:encodings}, <<\elpiIn{x\ f x}>> has not equal to 
%<<\elpiIn{lam x\ app[f, x]}>>, since, even though the two terms represent the same
%concept, the latter uses the abstraction and application of the OL.
%
%In particular, the node \elpiIn{abs} stands for the lambda abstraction of the ML.
% The
% node \elpiIn{val} contains terms of the object language. The node for variable
% application is not really necessary, since, using the syntax of the ML, it is
% defined by putting two terms of the ML side by side. For example, the terms 
% <<\elpiIn{x\ f x}>> becomes <<\elpiIn{abs x\ f x}>>.

The memory \Fo{} is called \elpiIn{fsubst} while the one of \Ho is
called \elpiIn{subst}.
Both have the invariant that they are not cyclic, TODO explain.

\subsection{equational theory and unification}

@@

\paragraph{Term dereferencing: $\rho s$ vs. $\sigma t$}

Since in our encoding we explicitly drag substitution mappings, we need to
dereference variables of the two languages. This is particularly useful to check
if two terms in the OL (resp. in the ML) are equal. The constructors
representing rigid terms, namely \elpiIn{app}, \elpiIn{lam} and \elpiIn{con},
make the dereferencing procedure to recurse over the their subterms. The code
below display this behavior for the dereferencing performed by the OL:

\begin{elpicode}
type fderef fsubst -> fm -> fm -> o.
fderef S (fapp [fuva N|A]) R :- set N S (fapp B),
  append A B C, fderef S (fapp C) R.
fderef S (fapp [fuva N|A]) (fapp [fuva N|B]) :- unset N S,
  forall2 (fderef S) A B.
fderef S (fapp A) (fapp B) :- forall2 (fderef S) A B.
fderef S (flam F) (flam G) :- 
  pi x\ fderef S x x => fderef S (F x) (G x).
fderef _ (fcon C) (fcon C).
fderef S (fuva N) R :- set N S T, fderef S T R.
fderef _ (fuva N) (fuva N) :- unset N S.
\end{elpicode}

More attention should be payed when the dereferencing encounters a subterm that
is unification variable. The following new rules should be added to the previous
code snipper to complete the dereferencing of variables of the OL:

A very similar implementation for the dereferencing performed by the ML, given 
in the appendix, can be encoded respecting the signature:

\marginpar{what about app uva?}
\begin{elpicode}
  type deref subst -> tm -> tm -> o.
  % ... the same
  deref S (uva N A) R :- set N S T,
    instantiate T A T', deref S T' R.
  deref S (uva N A) (uva N B) :-unset N S,
    forall2 (deref S) A B.
\end{elpicode}

In particular, 1) if the variable \elpiIn{fuva N} is assigned to a term
\elpiIn{T} in a substitution mapping \elpiIn{S}, we return the result of
dereferencing \elpiIn{T}; 2) if the current term is the application of a
variable \elpiIn{fuva N} assigned to a list of terms \elpiIn{L} and \elpiIn{uv N}
is assigned to a term \elpiIn{T}, we return the result of dereferencing the beta
reduction of \elpiIn{T} to \elpiIn{L}. This is due to the semantics of the
unification variables within the OL; 3) finally, if \elpiIn{uv N} is not
assigned, then the same variable is returned.

A different reasoning is to be addressed to the variables of the ML. Firstly, a
meta variable cannot appear in the \elpiIn{app} node as the first element of the
list, we will explain why in \cref{sec:compilation}; secondly, variables in the
ML have a scope which should be taken into account. The following few lines of
code outline the rules of dereferencing for the ML:

where, 1) if the variable \elpiIn{hov N L} has an assignment \elpiIn{T}, we
return the dereferencing of the beta reduction \elpiIn{T} to the scope
\elpiIn{L}.

\begin{figure}
  \begin{elpicode}
    ~\PYG{k+kd}{type} \PYG{n+nf}{(\Eo)} \PYG{k+kt}{ftm -> ftm -> o}~.
    % congruence
    fapp A ~\Eo~fapp B :- forall2 ~(\Eo)~ A B.
    flam F ~\Eo~flam G :- pi x\ x ~\Eo~x => F x ~\Eo~G x.
    fcon C ~\Eo~fcon C.
    fuva N ~\Eo~fuva N.
    % eta
    flam F ~\Eo~T :- pi x\ beta T [x] (T' x),                 ~($\customlabel{clause:eta1}{\eta_1}$)~
      x ~\Eo~x => F x ~\Eo~T' x.
    T ~\Eo~flam F :- pi x\ beta T [x] (T' x),                 ~($\customlabel{clause:eta2}{\eta_2}$)~
      x ~\Eo~x => T' x ~\Eo~F x.
    % beta
    fapp [flam X | TL] ~\Eo~T :- beta (flam X) TL T', T' ~\Eo~T. ~($\customlabel{clause:beta1}{\beta_1}$)~
    T ~\Eo~fapp [flam X | TL] :- beta (flam X) TL T', T ~\Eo~T'. ~($\customlabel{clause:beta2}{\beta_2}$)~
  \end{elpicode}
  \caption{Equal predicate OL}
  \label{code:eq-pred}
  \Description[code:equal-pred]{code:equal-pred}
\end{figure}

\begin{figure}
  \begin{elpicode}
    ~\PYG{k+kd}{type} \PYG{n+nf}{(\Ee)} \PYG{k+kt}{tm -> tm -> o}~.
    app A ~\Ee~fapp B :- forall2 ~(\Ee)~ A B.
    lam F ~\Ee~flam G :- pi x\ x ~\Ee~x => F x ~\Ee~G x.
    con C ~\Ee~fcon C.
    uva N A ~\Ee~fuva N B :- forall2 ~(\Ee)~ A B.
  \end{elpicode}
  \caption{Equal predicate ML}
  \label{code:eq-pred-meta}
  \Description[code:equal-pred-meta]{code:equal-pred-meta}
\end{figure}

@@

\begin{figure}
\begin{elpicode}
~\PYG{k+kd}{type} \PYG{n+nf}{(\Ue)} \PYG{k+kt}{tm -> tm -> subst -> subst o}~.
% Congruence
(app A ~\Ue~app B) S S1 :- fold2 ~\Ue~A B S S1.
(lam F ~\Ue~lam G) S S1 :- 
  pi x\ (pi S\ (x ~\Ue~x) S S) => (F x ~\Ue~G x) S S1.
~\Ue~(con C) (con C) S S.
% deref
(uva N A ~\Ue~T) S S1 :- set N S F,
  beta-abs F A T1, (T1 ~\Ue~T) S S1.
(T ~\Ue~uva N A) S S1 :- set N S F,
  beta-abs F A T1, (T ~\Ue~T1) S S1.
% flex-flex
(uva N A ~\Ue~uva M B) S S3 :-  unset M, unset N,
  pattern-fragment A, pattern-fragment B, 
  new S W S1, prune W Args1 B Ass, 
  assign N S1 Ass S2, assign M S2 Ass S3.
% assignment
(uva N A ~\Ue~T) S S1 :- pattern-fragment A,
  not (T = uva _ _), not_occ N S T,  
  bind A T T1, assign N S T1 S1.
(T ~\Ue~uva N A) S S1 :- pattern-fragment A,
  not (T = uva _ _), not_occ N S T, 
    bind A T T1, assign N S T1 S1.
%  Note: (a infix b) c d === (infix) a b c d

\end{elpicode}
\caption{The \Ue predicate // TODO move to the appendix}
\label{code:ue-pred}
\Description[code:unif-pred]{code:unif-pred}
\end{figure}

On the other hand, unification in the ML needs to be defined. In
\cref{code:ue-pred}, we give an implementation of \Ue. This predicate has four
arguments, the two terms $t_1$ and $t_2$ to unify, the old substitution map
$\rho_1$, and the new substitution map $\rho_2$, with the invariant $\rho_1
\subseteq \rho_2$. The first three rules unify terms with same rigid heads, and
call the unification relation on the sub-terms. If $t_1$ (resp. $t_2$) is an
assigned variables, $t_1$ is dereferenced to $t_1'$ (resp. $t_2'$) and the
unification is called between $t_1'$ and $t_2$ (resp. $t_1$ and $t_2'$). If both
terms are unification variables, we test that their arguments are in the pattern
fragment, we allocate a new variable $w$ in $\rho_1$ such that $w$ is the
pruning of the arguments of $t_1$ and $t_2$, we assign both $t_1$ and $t_2$ to
$w$ and return the new mapping $\rho_2$ containing all the new variable
assignment. Finally, if only one of the two terms is an unification variable
$v$, after having verified that $v$ does not occur in the other term $t$, we
bind $v$ to $t$ and return the new substitution mapping.

@@

>>>>

\section{Preliminaries: \Fo{} and \Ho}
\label{sec:lang-spec}

@@

\begin{figure}[H]
  \begin{tabular}{ll}
  \begin{minipage}{0.21\textwidth}
  \begin{elpicodetab}
kind fm type.
type fapp list fm -> fm.
type flam (fm -> fm) -> fm.
type fcon string -> fm.
type fuva nat -> fm.
\end{elpicodetab}
  \end{minipage}
  &
  \begin{minipage}{0.24\textwidth}
  \begin{elpicodetab}
kind tm type.
type app list tm -> tm.
type lam (tm -> tm) -> tm.
type con string -> tm.
type uva nat -> list tm -> tm.
  \end{elpicodetab}
  \end{minipage}
  \end{tabular}
  \caption{\Fo{} and \Ho language}\vspace{0.3em}
  \label{code:common-terms}
  \Description[code:common-terms]{code:common-terms}
\end{figure}

\noindent
In the case of \Fo{} unification variables \elpiIn{fuva} have no explicit scope:
Unification variables standing for functions are applied their arguments
via the \elpiIn{fapp} constructor.
For example in the statement of the instance \coqIn{forall_dec} the
term \coqIn{P x} is represented as \elpiIn{fapp[fuva N, x]}, where \elpiIn{N}
is of type \elpiIn{nat} and \elpiIn{x} is a bound variable.

@@

\noindent
The \elpiIn{name} builtin predicate tests if a term is a bound variable. The
compiler~\ref{sec:xxx} needs to support terms outside \llambda for practical
reasons, so we don't assume all out terms are in \llambda but rather test.

@@

\begin{elpicode}
typeabbrev memory A (list (option A)).
type set?   nat -> memory A -> A -> o.
type unset  nat -> memory A -> o.
type assign nat -> memory A -> A -> memory A -> o.
\end{elpicode}

\noindent 
If a memory cell is \elpiIn{none}, then the corresponding unification variable
is not set. \elpiIn{assign} sets an unset cell to the given value.

Since in \Ho unification variables have a scope, their solution needs to be
abstracted over it in order to enable the instantiation of a single
solution to different scopes. On the contrary solutions to \Fo variables
are plain terms.

\begin{elpicode}
  typeabbrev fsubst (memory ftm).
  kind assmt type.
  type abs (tm -> assmt) -> assmt.
  type val tm -> assmt.
  typeabbrev subst (memory assmt).
\end{elpicode}

\noindent
We call \elpiIn{fsubst} the memory of \Fo{}, while we call \elpiIn{subst}
the one of \Ho.
Both have the invariant that they are not cyclic, TODO explain.

\subsection{Equational theory and Unification}

@@

\paragraph{Term dereferencing: $\rho s$ and $\sigma t$}

Since in our encoding we explicitly carry a substitution we need to
define the operation that applies it to a term. Its definition
takes care to normalize (flatten) applications, for example
it turns
\elpiIn{(app [app [con"f",con"a"],con"b"])}
into
\elpiIn{(app [con"f",con"a",con"b"])}.

dereference variables of the two languages. This is particularly useful to check
if two terms in the OL (resp. in the ML) are equal. The constructors
representing rigid terms, namely \elpiIn{app}, \elpiIn{lam} and \elpiIn{con},
make the dereferencing procedure to recurse over the their subterms. The code
below display this behavior for the dereferencing performed by the OL:

\begin{elpicode}
type fderef fsubst -> fm -> fm -> o.
% flattening
fderef S (fapp [fuva N|A]) R :- set? N S (fapp B), !,
  append A B C, fderef S (fapp C) R.
% traversal
fderef S (fapp A) (fapp B) :- map (fderef S) A B.
fderef S (flam F) (flam G) :- 
  pi x\ fderef S x x => fderef S (F x) (G x).
fderef _ (fcon C) (fcon C).
% dereferencing
fderef S (fuva N) R :- set? N S T, fderef S T R.
fderef _ (fuva N) (fuva N) :- unset N S.
\end{elpicode}

\noindent
We use the cut operator to keep the code compact. It is possible to rewrite
the rule for application traversal so that it is mutually exclusive with the
first one, but that requires a rather verbose analysis of the head of \elpiIn{A}.

The corresponding code for \Ho is similar, we only show the last two rules that
differ in a substantial way:

\begin{elpicode}
type deref subst -> tm -> tm -> o.
% ... similar to above ...
deref S (uva N A) R :- set? N S F, move F A T, deref S T R.
deref S (uva N A) (uva N B) :- unset N S, map (deref S) A B.
type move assignment -> list tm -> tm -> o.
move (abs F) [X | L] R :- move (F X) L R.
move (val A)         [] A :- !.
move (val (app A))   L (app X) :- append A L X.
move (val (uva N A)) L (uva N X) :- append A L X.
move (val (con H))   L (app [con H | L]).
\end{elpicode}

\noindent
Note that when the substitution \elpiIn{S} maps a unification variable
\elpiIn{N} to an assignment \elpiIn{F} we ....

Important!!!
A different reasoning is to be addressed to the variables of the ML. Firstly, a
meta variable cannot appear in the \elpiIn{app} node as the first element of the
list, we will explain why in \cref{sec:compilation}

\begin{figure}
  \begin{elpicode}
    ~\PYG{k+kd}{type} \PYG{n+nf}{(\Eo)} \PYG{k+kt}{ftm -> ftm -> o}~.
    % congruence
    fapp A ~\Eo~fapp B :- map ~(\Eo)~ A B.
    flam F ~\Eo~flam G :- pi x\ x ~\Eo~x => F x ~\Eo~G x.
    fcon C ~\Eo~fcon C.
    fuva N ~\Eo~fuva N.
    % eta
    flam F ~\Eo~T :- pi x\ beta T [x] (T' x),                 ~($\customlabel{clause:eta1}{\eta_1}$)~
      x ~\Eo~x => F x ~\Eo~T' x.
    T ~\Eo~flam F :- pi x\ beta T [x] (T' x),                 ~($\customlabel{clause:eta2}{\eta_2}$)~
      x ~\Eo~x => T' x ~\Eo~F x.
    % beta
    fapp [flam X | TL] ~\Eo~T :- beta (flam X) TL T', T' ~\Eo~T. ~($\customlabel{clause:beta1}{\beta_1}$)~
    T ~\Eo~fapp [flam X | TL] :- beta (flam X) TL T', T ~\Eo~T'. ~($\customlabel{clause:beta2}{\beta_2}$)~
  \end{elpicode}
  \caption{Equal predicate OL}
  \label{code:eq-pred}
  \Description[code:equal-pred]{code:equal-pred}
\end{figure}

\begin{figure}
  \begin{elpicode}
    ~\PYG{k+kd}{type} \PYG{n+nf}{(\Ee)} \PYG{k+kt}{tm -> tm -> o}~.
    app A ~\Ee~fapp B :- map ~(\Ee)~ A B.
    lam F ~\Ee~flam G :- pi x\ x ~\Ee~x => F x ~\Ee~G x.
    con C ~\Ee~fcon C.
    uva N A ~\Ee~fuva N B :- map ~(\Ee)~ A B.
  \end{elpicode}
  \caption{Equal predicate ML}
  \label{code:eq-pred-meta}
  \Description[code:equal-pred-meta]{code:equal-pred-meta}
\end{figure}

@@

\begin{elpicode}
~\PYG{k+kd}{type} \PYG{n+nf}{(\Ue)} \PYG{k+kt}{tm -> tm -> subst -> subst -> o}~.
\end{elpicode}

On the other hand, unification in the ML needs to be defined. In
\cref{code:ue-pred}, we give an implementation of \Ue but
that is actually what our meta language provides as a builtin.

This predicate has four
arguments, the two terms $t_1$ and $t_2$ to unify, the old substitution map
$\rho_1$, and the new substitution map $\rho_2$, with the invariant $\rho_1
\subseteq \rho_2$. The first three rules unify terms with same rigid heads, and
call the unification relation on the sub-terms. If $t_1$ (resp. $t_2$) is an
assigned variables, $t_1$ is dereferenced to $t_1'$ (resp. $t_2'$) and the
unification is called between $t_1'$ and $t_2$ (resp. $t_1$ and $t_2'$). If both
terms are unification variables, we test that their arguments are in the pattern
fragment, we allocate a new variable $w$ in $\rho_1$ such that $w$ is the
pruning of the arguments of $t_1$ and $t_2$, we assign both $t_1$ and $t_2$ to
$w$ and return the new mapping $\rho_2$ containing all the new variable
assignment. Finally, if only one of the two terms is an unification variable
$v$, after having verified that $v$ does not occur in the other term $t$, we
bind $v$ to $t$ and return the new substitution mapping.

@@

\clearpage
\section*{Appendix}

Note that \elpiIn{(a infix b) c d} de-sugars to \elpiIn{(infix) a b c d}.

\begin{figure*}
\begin{elpicode}
~\PYG{k+kd}{type} \PYG{n+nf}{(\Ue)} \PYG{k+kt}{tm -> tm -> subst -> subst -> o}~.
% Congruence
(app A ~\Ue~app B) S S1 :- fold2 ~(\Ue)~A B S S1.
(lam F ~\Ue~lam G) S S1 :- pi x\ (pi S\ (x ~\Ue~x) S S) => (F x ~\Ue~G x) S S1.
~\Ue~(con C) (con C) S S.
% deref
(uva N A ~\Ue~T) S S1 :- set? N S F, move F A T1, (T1 ~\Ue~T) S S1.
(T ~\Ue~uva N A) S S1 :- set? N S F, move F A T1, (T ~\Ue~T1) S S1.
% flex-flex
(uva N A ~\Ue~uva M B) S S3 :-  unset M, unset N,
  distinct-names A, distinct-names B, 
  new S W S1, prune W Args1 B Ass, 
  assign N S1 Ass S2, assign M S2 Ass S3.
% assignment
(uva N A ~\Ue~T) S S1 :- distinct-names A, not (T = uva _ _), not_occ N S T,  
  bind A T T1, assign N S T1 S1.
(T ~\Ue~uva N A) S S1 :- distinct-names A, not (T = uva _ _), not_occ N S T, 
  bind A T T1, assign N S T1 S1.

type distinct-names list tm -> o.
distinct-names [].
distinct-names [X|XS] :- name X, not(mem X XS),
distinct-names XS.

typeabbrev memory A (list (option A)).
type set? nat -> memory A -> A -> o.
set? N S T :- nth N S (some T).
type unset nat -> memory A -> o.
unset N S :- nth N S none.
type assign nat -> memory A -> A -> memory A -> o.
assign z [none|M] T [some T|M].
assign (s N) [X|M] T [X|M1] :- assign N M T M1. 
kind nat type.
type z nat.
type s nat -> nat.
type nth nat -> list A -> A -> o.
nth z [X|_] X.
nth (s N) [_|L] X :- nth N L X.

type new memory A -> nat -> memory A -> o.
type prune .
type move .
type beta.
TODO

\end{elpicode}
\caption{Implementation of the \Ue predicate for \Ho}
\label{code:ue-pred}
\Description[code:unif-pred]{code:unif-pred}
\end{figure*}
